Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-03-30 bbbq_article.tex
Processed feedback Geert 2021-03-24, fix #196

\richel{I wonder if the order of the Methods, should match the improved
order of the Results as well, so first the MHC-I and MHC-II presentation,
followed by the experimental confirmation}

@@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{ncbi_peregrine_results/fig_conservation.png}
    \label{fig:conservation}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{ncbi_peregrine_results/fig_f_snps_found_and_expected.png}
    \label{fig:f_snps_found_and_expected}
  \end{subfigure}
  \caption{
      \textbf{(A)} 
      The number of SNPs in TMHs as expected by chance (left bar) 
      and found in the dbSNP database (right bar).
      Percentages are added to show the relative conservation
      of SNPs in TMHs found.
      \textbf{(B)}
      Percentage of SNPs found in TMHs.
      Each point shows the predicted percentage of
      TMH (x-axis) and the observed occurrence of SNPs being located
      within a TMH (y-axis) for a protein.
      The dashed diagonal line shows the line of equality (i.e.
      equal conservation of TMHs and soluble protein regions).
      The blue line is a linear trend line that includes only the
      transmembrane proteins.
  }
\end{figure}

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.3\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{ncbi_peregrine_results/fig_conservation_per_spanner.png}
    \label{fig:conservation_per_spanner}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.55\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{ncbi_peregrine_results/fig_f_snps_found_and_expected_per_spanner.png}
    \label{fig:f_snps_found_and_expected_per_spanner}
  \end{subfigure}
  \caption{
      \textbf{(A)} 
      The number of SNPs in TMHs as expected by chance (red bars) 
      and found in the dbSNP database (right bars), 
      after splitting the data into single- and multi-spanners.
      Percentages are added to show the relative conservation
      of SNPs in TMHs found.
      \textbf{(B)} 
      Percentage of SNPs found in TMPs predicted to have only a single
      (left) or multiple (right) TMHs.
      The dashed diagonal lines show the expected line of equality 
      (i.e., equal conservation of TMHs and solvent-exposed 
      protein regions).
  }
\end{figure}

@@

Per protein, we calculated two percentages: 
(1) the percentage of the total protein predicted to be  TMHs, 
and (2) the percentage of SNPs located within these predicted TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
We performed a linear regression analysis on the TMPs
and added a 95\% confidence interval on this regression.
Would the confidence interval of the regression line 
overlap with the line of equality,
this would indicate that SNPs occur just as likely within TMHs 
and solvent-exposed protein domains (i.e., no higher conservation of TMHs). However, we observed that ...[GEERT give brief explanation in words of what is seen in the graph]
This finding suggests that SNPs 
occur more often in soluble domains than in TMHs.

To statistically test whether SNPs occur 
just as likely in soluble domains as
in TMHs, we first calculated the number of SNPs that are expected to be
found in TMHs of TMPs by chance ($E(n_{SNP\_TMH})= n_{SNP} \cdot f_{TMH} \approx 4,056$), by multiplying the number of SNPs predicted to be in TMHs ($n_{SNP} = 21,738$ ) with the fraction of the TMPs that are predicted TMH. The actual number of SNPs predicted to be located within TMHs
is $n_{SNP\_TMH} = 3,903$, which is lower than the expectation
of $E(n_{SNP\_TMH}) \approx 4,056$.

To test whether this difference is significant,
we used a binomial Poisson distribution, where the number of trails ($n$) equals the number of SNPs, which is 21,738. The probability of success
for the $i$th TMP ($p\_i$), is the percentage of each TMP that is TMH.
We calculated the probability of having 3,903 or less successes.
We used the type I error cut-off value of $\alpha = 2.5\%$.
The chance to find, within TMHs, this amount or less SNPs 
equals [RJCB: recalculate]
$1.6201 \cdot 10^{-9}$.

Using a alpha value of $0.05$, as is common practice for an uninformed
expectation, we conclude that there are significantly less SNPs
in TMHs as expected by chance.
See figure \ref{fig:f_tmh_ncbi} for the percentages of TMHs in the transmembrane proteins.[GEERT: staat deze figuur hier wel goed? Niet naar boven?] The effect size is that, 
when randomly selecting SNPs,
per 100 SNPs found in soluble protein domains, 
one finds 91 SNPs in TMHs,
as depicted in figure \ref{fig:conservation}. [GEERT: hieronder had je indentieke tekst, maar verwees naar andere figuur \ref{fig:conservation_per_spanner}. Ik denk dat je deze hieronder kan samenvoegen]

We splitted this analysis based on the number of TMHs
of each TMP. 
We hypothesized that single-membrane spanners (i.e., proteins
with a single predicted TMH) are less conserved than multi-membrane spanners (with $>$1 TMH),
because multi-membrane spanners
might have protein-protein interactions between their TMHs, 
for example to accommodate active sites, and 
thus might have additional structural constraints.
As can be seen in figure \ref{fig:f_snps_found_and_expected_per_spanner}, 
both single- and multi-spanners are evolutionarily conserved.[GEERT: maar een sterker dan de andere? Kun je die conclusie niet trekken?]

@@

Not only do we predict that MHC-I and MHC-II TMH-derived epitopes 
are presented , we conclude that these are additionally
presented more often than expected by chance 
alone (figure \ref{fig:bbbq_1_smart_results}).
This in silico finding is a general pattern 
that persists in both MHC-I and MHC-II, 
for humans, bacteria and viruses.

% \paragraph{MHC-II is expected to present TMHs}
\frans{we moeten iets schrijven over de verschillende haplotypen wat is the occurence van de haplotypen met andere worden hoe relevant zijn de haplotypen die veel TMH presenteren?}
The over-presentation of TMHs for most MHC-II haplotypes 
is more pronounced than for MHC-I haplotypes,
as up to a third of the predicted MHC-II binders 
\richel{Hier ben ik nu met verwerken...} 
from the human and MTb proteome stem from TMHs. \frans{ik zoek nog naar associaties} 
such as \verb;HLA-DQA1*0101/DQB1*0501;

This results from this study support our hypothesis that TMH-derived epitopes are over-presented
in MHC-II, similarly to we previously showed for MHC-I [CITE BIANCHI].[GEERT: deel gewist, want was letterlijke herhaling van hierboven. Beter is zeggen dat (1) MHC-II presentatie belangrijk is voor B cell activatie. (2) B cellen antilichamen produceren, en dan antilichamen solvent-exposed protein regions herkennen. (3) Dat onze analyse suggereert HMC-II presentatie voornamelijk andere (TMH of buried) regio's betreft. (4) Dat T helper cellen dus andere delen van antigene eiwitten herkennen (MHC-II gepresenteerd door de B cellen) dan de BCR bindt.]

@@

\geert{
  Also compare your findings with previous efforts 
  to look at conservation of TMHs
}

@@

Figure \ref{fig:f_tmh_ncbi} shows the distribution 
of the percentages of TMH of transmembrane proteins.

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_f_tmh_ncbi.png}
  \caption{
    Distribution of the percentages of TMH of transmembrane proteins.
    Dashed vertical line: mean \% TMH in all proteins.
    Dotted vertical line: mean \% TMH in TMPs.
  }
  \label{fig:f_tmh_ncbi}
\end{figure}

>>>>

@@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{ncbi_peregrine_results/fig_conservation.png}
    \label{fig:conservation}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{ncbi_peregrine_results/fig_f_snps_found_and_expected.png}
    \label{fig:f_snps_found_and_expected}
  \end{subfigure}

   \vfill

  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{ncbi_peregrine_results/fig_f_tmh_ncbi.png}
    \label{fig:f_tmh_ncbi}
  \end{subfigure}

  \caption{
    \textbf{(A)} 
    The number of SNPs in TMHs as expected by chance (left bar) 
    and found in the dbSNP database (right bar).
    Percentages are added to show the relative conservation
    of SNPs in TMHs found.
    \textbf{(B)}
    Percentage of SNPs found in TMHs.
    Each point shows the predicted percentage of
    TMH (x-axis) and the observed occurrence of SNPs being located
    within a TMH (y-axis) for a protein.
    The dashed diagonal line shows the line of equality (i.e.
    equal conservation of TMHs and soluble protein regions).
    The blue line is a linear trend line that includes only the
    transmembrane proteins.
    \textbf{(C)}
    Distribution of the percentages of TMH of TMPs.
  }
\end{figure}

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{ncbi_peregrine_results/fig_conservation_per_spanner.png}
    \label{fig:conservation_per_spanner}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{ncbi_peregrine_results/fig_f_tmh_ncbi.png}
    \label{fig:f_tmh_ncbi_per_spanner}
  \end{subfigure}

  \vfill

  \begin{subfigure}[t]{\textwidth}
    \centering
    \caption{}
    \includegraphics[width=\linewidth]{ncbi_peregrine_results/fig_f_snps_found_and_expected_per_spanner.png}
    \label{fig:f_snps_found_and_expected_per_spanner}
  \end{subfigure}
  \caption{
      \textbf{(A)} 
      The number of SNPs in TMHs as expected by chance (red bars) 
      and found in the dbSNP database (right bars), 
      after splitting the data into single- and multi-spanners.
      Percentages are added to show the relative conservation
      of SNPs in TMHs found.
      \textbf{(B)} 
      \richel{THIS IS A PLACEHOLDER}
      Distribution of the proportion of amino acids residing
      in the plasma membrane.
      \textbf{(C)} 
      Percentage of SNPs found in TMPs predicted to have only a single
      (left) or multiple (right) TMHs.
      The dashed diagonal lines show the expected line of equality 
      (i.e., equal conservation of TMHs and solvent-exposed 
      protein regions).
  }
\end{figure}

@@

Per protein, we calculated two percentages: 
(1) the percentage of the total protein predicted to be TMHs, 
and (2) the percentage of SNPs located within these predicted TMHs.
Each percentage pair was plotted in figure \ref{fig:f_snps_found_and_expected}.
Most our TMPs had less than half their residues located in the 
plasma membrane.
The proportion of SNPs found in TMHs varied from 
none (i.e. all SNPs were in
soluble regions) to all.
To determine if it is likelier to find SNPs within either TMHs or
soluble-facing domains, we performed a linear regression analysis,
and added a 95\% confidence interval on this regression.
This linear fit nearly goes through the origin and has a slope
below the line of equality,
which shows that less SNPs are found in TMHs than expected by chance.

We determined the probability to find the observed amount
of SNPs in TMHs by chance, i.e. when assuming SNPs occur 
just as likely in soluble domains as in TMHs.
We used a binomial Poisson distribution, 
where the number of trails ($n$) equals the number of SNPs, 
which is 21,208. 
The probability of success for the $i$th TMP ($p\_i$), 
is the percentage of residues within a TMH per TMP. 
These probabilities are shown as a histogram 
in figure \ref{fig:f_tmh_ncbi}). 
The expected number of SNPs expected to be found in 
TMHs by chance equals $\sum{p} \approx 4,141$.
As we observed 3,903 SNPs in TMHs, 
we calculated the probability of having that amount or less successes.
We used the type I error cut-off value of $\alpha = 2.5\%$.
The chance to find, within TMHs, this amount or less SNPs 
equals $6.8208 \cdot 10^{-11}$.
We determined the relevance of this finding, by
calculating how much less SNPs are found in TMHs,
when compared to soluble regions, which is the
ratio between the number of SNPs found in TMHs
versus the number of SNPs as expected by chance.
In effect, per 1000 SNPs found in soluble protein domains, 
one finds 918 SNPs in TMHs,
as depicted in figure \ref{fig:conservation}.

We split this analysis based on the number of TMHs
of each TMP. 
We hypothesized that single-membrane spanners (i.e., proteins
with a single predicted TMH) are less conserved than multi-membrane spanners (with $>$1 TMH),
because multi-membrane spanners
might have protein-protein interactions between their TMHs, 
for example to accommodate active sites, and 
thus might have additional structural constraints.
From the split data, we did the same analysis:
figure \ref{fig:f_snps_found_and_expected_per_spanner} 
shows, the percentage a protein resides in the membrane versus the
percentage of SNPs found in those TMH regions, per spanner.
In both cases, a linear regression shows that less
SNPs are found in TMHs, than expected by chance.

Also per spanner, we determined the probability to find the 
observed amount of SNPs in TMHs by chance.
For single-spanners, we found 452 SNPs in TMH, where
$\approx462$ were expected by chance. 
The chance to observe this or a lower number by chance is 
$0.319$. As this chance was higher than our $\alpha = 0.025$
and we considered this no significant effect.
For the single-spanners, we found 3,351 SNPs in TMH, where
$\approx3678$ were expected by chance. 
The chance to observe this or a lower number by chance is 
$8.315841 \cdot 10^{-12}$, 
which means we attribute the lower number of SNPs found
to an effect, instead of luck.

Also, per spanner type, 
we determined the relevance of this finding, by
calculating how much less SNPs are found in TMHs,
when compared to soluble regions,
as depicted in figure \ref{fig:conservation_per_spanner}.
In effect, per 1000 SNPs found in soluble protein domains, 
one finds 978 SNPs in TMHs 
in single spanners (although this is not significant)
and 911 SNPs in TMHs in multi-spanners.

@@

Not only do we predict that MHC-I and MHC-II TMH-derived epitopes 
are presented, we conclude that these are additionally
presented more often than expected by chance 
alone (figure \ref{fig:bbbq_1_smart_results}).
This in silico finding is a general pattern 
that persists in both MHC-I and MHC-II, 
for humans, bacteria and viruses.

% \paragraph{MHC-II is expected to present TMHs}
The results from this study support our hypothesis 
that TMH-derived epitopes are over-presented
in MHC-II, similarly to we previously showed for MHC-I (\cite{bianchi2017}). 
The over-presentation of TMHs for most MHC-II haplotypes 
appears to follow a similar pattern as MHC-I haplotypes,
although MHC-II presents slightly more TMH-derived
epitopes (see Supplementary Materials, figure \ref{fig:rel_presentation}
for a direct comparison). 
Unlike MHC-I, the human proteome over-presents TMH-derived
epitopes for all MHC-II haplotypes that we have investigated.
Possible there are MHC-II haplotypes that present less TMH-derived
epitopes than expected by chance.
The haplotype \verb;HLA-DQA1*0101/DQB1*0501; had the highest
over-presentation of TMH-derived epitopes,
which we suggest is the haplotype of choice 
to investigate the intracellular mechanism of TMH-derived
antigen presentation.

\richel{Geert, ik kan je hulp goed gebruiken bij dit stukje}
Presentation of antigens by MHC-II is important for the activation of
B cells \richel{je bedoelt denk ik T cellen hier?}. 
B cell produce antibodies that recognize only solvent-exposed protein regions,
thus the B cell receptor is inherently unable to present hydrophobic regions.
Our analysis suggests, however, that MHC-II
over-presents TMH regions. A helper T cell thus needs an 
\geert{
  Beter is zeggen dat (1) MHC-II presentatie belangrijk is voor B cell activatie. (2) B cellen antilichamen produceren, en dan antilichamen solvent-exposed protein regions herkennen. (3) Dat onze analyse suggereert HMC-II presentatie voornamelijk andere (TMH of buried) regio's betreft. (4) Dat T helper cellen dus andere delen van antigene eiwitten herkennen (MHC-II gepresenteerd door de B cellen) dan de BCR bindt.
}

@@

There have been more efforts to look at the conservation of TMHs,
using different methodologies.
One such example is \ref{stevens2001substitution}, 
in which aligned protein sequence data was used.
Also this study found that TMHs are evolutionarily more conserved,
as the mean amino acid substitution rate in TMHs is about ten
percent lower,
which is the same value as this study found.
Another example is a study that estimates the conservation
scores for TMH and soluble regions based on 
evolutionary related protein alignments,
and found that TMHs are more conserved, 
with a conservation score that was 17\% higher in 
TMHs (\ref{oberai2009structural}).
Note that the last study also used human SNPs
and found that mutations in TMH are likelier to cause
a disease, again underlining the importance of TMHs.

@@

%Figure \ref{fig:f_tmh_ncbi} shows the distribution 
%of the percentages of TMH of transmembrane proteins.
%
%\begin{figure}[!htbp]
%  \includegraphics[width=\textwidth]{ncbi_peregrine_results/fig_f_tmh_ncbi.png}
%  \caption{
%    Distribution of the percentages of TMH of transmembrane proteins.
%    Dashed vertical line: mean \% TMH in all proteins.
%    Dotted vertical line: mean \% TMH in TMPs.
%  }
%  \label{fig:f_tmh_ncbi}
%\end{figure}
